---
title: "Lake Ice Phenology"
author: 'Quinn Thomas'
output:
  github_document: default
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Data

The data for this module are provided within the GitHub repository in the `data`
sub-directory.

-   Data: `data/lake_ice_off_data_2022.xlsx`

# Import and Prep

The data are in a multi-sheet excel file named
`data/lake_ice_off_data_2022.xlsx`. You will need to use the `read_excel()`
function in the `readxl` package to import the data into R as a data frame. You
can learn more about importing from excel
[here](https://r4ds.hadley.nz/spreadsheets.html#prerequisites)

```{r}

```

Now look at a preview of the data using the `head()` function. The `doy` in
`ice_off_doy` stands for "day of year" and is the day within the year where ice
off occurred (January 1 = doy 1)

```{r}

```

There should be 3 columns and \~2,659 rows.

## Tidy the data

We will practice key skills in the module using a focal lake. Use the `filter`
function to filter the larger data to only contain the data from Lake Sunapee, a
lake in New Hampshire. Filter the data `ice_data` by the variable `lake_name` to
create a data frame called `sunapee`.

```{r}

```

You will be using the `sunapee` data for the rest of the module until Question 11

## Visualize the data

**Question 1:**

To visualize the data from Lake Sunapee, create an x-y plot with the year
(`year`) on the x-axis and the day of year (`ice_off_doy`) on the y-axis. Plot
the points (`geom_point`) and connect the points with lines (`geom_line`). Be
sure that the title, x-axis and y-axis labels are clearly labeled.

**Answer 1:**

```{r}

```

# How is ice cover changing over time at Lake Sunapee?

We want to model how the ice-off date is changing through time. Our approach in
this module is to use a linear regression that predicts the ice-off day of year
as a function of the year. The `lm` function will estimate the parameters of the
DOY = b + m \* YEAR linear regression where b is the intercept and m is the
slope.

**Question 2:**

What do you expect the sign (plus, minus, or no change) to be for the slope of
the relationship? Why?

**Answer 2:**

Use the `lm` function to predict `ice_off_doy` as a function (`~`) of the year
for the data.frame called `sunapee`. You can learn more about the lm function [here](https://www.datacamp.com/tutorial/linear-regression-R)

Create an object called `fit` that contains results of the linear model.

```{r}

```

Then use the function `summary` to explore the result of the `lm` function

```{r}

```

The results of your linear model are in an object named `fit`, so you can find
the regression coefficients at `fit$coefficients[1]` (the intercept) and
`fit$coefficients[2]` (the slope). Confirm that `fit$coefficients[1]` and
`fit$coefficients[2]` match the values expected from the summary command you ran
above.

```{r}
fit$coefficients[1]
fit$coefficients[2]
```

**Question 3:**

What is the direction of the slope? How does this match your expectation from
Question 2?

**Answer 3:**

# Is our model any good?

## R2

First, it is important to understand how much variation is explained by the
model using the R2 value in the `summary`. A value of R2 close to 1 means that
the predictor (`year`) explains nearly all of the variation in the variable we are predicting: `ice_off_doy`. A value of R2 close to 0 explains very little of the
variation in `ice_off_doy`. Since we want to be able to explain how
`ice_off_doy` is changing over time using `year,` we are aiming for a high R2. A
low R2 would indicate that our model isn't particular good at explaining what
future values of `ice_off_doy` will be.

**Question 4:**

What is the R2 for the regression? Based on the R2, is `year` and `ice_off_doy`
strongly related?

**Answer 4:**

## Residuals normally distributed

Second, it is important that the residuals from our linear model are normally
distributed.  The following is a number distribution that you can visually compare to:

```{r}
ggplot(data = tibble(value = rnorm(1000, 0, 1))) +
  geom_histogram(aes(x = value), bins = 20)
```

We will use a histogram to visually assess whether this requirement is met.
First, you will need to calculate the residuals. The residuals are the
difference between the ice-off DOY that your linear regression predicts and the
observed DOY.

-   Residuals that do not average to zero indicate that your linear regression
    is biased.

-   Residuals that are widely spread out indicate that your linear regression is
    not very useful for prediction.

You will use a histogram to examine the residuals. First, the code uses the
coefficients to predict the ice-off DOY at each year. Second, calculate the
residuals by subtracting the predicted ice-off DOY from the observed DOY.

```{r}
sunapee <- sunapee %>% 
  mutate(predicted = fit$coefficients[1] + year * fit$coefficients[2], 
         residual = ice_off_doy - predicted) #Observed - Predicted
```

**Question 5:**

Use the [`add_predictions()`](https://modelr.tidyverse.org/reference/add_predictions.html) and [`add_residuals()`](https://modelr.tidyverse.org/reference/add_residuals.html) functions from the `modelr`
package (installs with the tidyverse but you need to load it with ` library()`) to add the predictions and residuals. This
accomplishes the same thing above but doesn't require you to specifically write
our your model with the coefficients (thus is more scalable to other models)

**Answer 5:**

```{r}

```

Now to visualize the residuals use `geom_histogram()` to plot them as a
histogram. I recommend using a bin width of 5 in the histogram.

**Question 6:**

Show Histogram. Do the residuals meet the assumptions of the linear regression? (are the
residuals normally distributed with the distribution centered around zero)?

**Answer 6:**

```{r}

```


## The slope is different from zero

A slope could appear to be higher or lower than zero but the uncertainty in the
slope is large enough that there is a chance that the value is 0 (thus limiting
our capacity believe the value for the slope is not zero). Here we will focus on
the p-value of the slope. The p value is the probability of obtaining a slope at
least as extreme as the one calculated from the data, assuming the truth of the
null hypothesis that the slope is 0 (from
<https://en.wikipedia.org/wiki/P-value>) In this case, what is the probability
of observing our slope value assuming the actually slope is 0. Smaller
probabilities mean there is lower change that our slope could actually be the
same as 0.

**Question 7:**

Based on the `summary()` above, what is the p-value for the slope?

**Answer 7:**

## The slope physically meaningful

A slope value could have a low probability of being 0 (those we have strong belief that it is not zero) but the magnitude of
change is functionally meaningless in the physical world (i.e., the rate of
change is so small that we would not think much of it). This can occur
particularly when analyzing large datasets.

**Question 8:**

Based on the slope value, how many more years are needed to pass before the lake
has 1 week of change in ice off? Do you think this is meaningful? Why?

**Answer 8:**

## Overall model assessment

**Question 9:**

How comfortable are you proceeding with using this linear model for additional
analysis and prediction? Why?

**Answer 9:**

Now plot the data for Lake Sunapee (just like above - year on x-axis vs. ice of doy on y axis) but
add the line from your regression results (e.g., fit intercept and slope). You will use the [`geom_abline` geom](https://ggplot2.tidyverse.org/reference/geom_abline.html). 

```{r}
#INSERT CODE
```

**Question 10:**
Use the regression model to predict the ice-off day of year in 2030.
Remember that year in the linear model (y = b + mx).

**Answer 10:**

```{r}
#INSERT CODE
```

# Is the ice-off day of year changing more quickly in more recent years at Lake Sunapee?

To answer this question, you need to divide the Sunapee dataset into two
datasets that represent an early and later period. For this analysis use the
year 1970 as a cut-off (one dataset is before 1970 and the other is 1970 and
after).

**Question 11:**

Use the skills developed above to evaluate the question "Is the ice-off day of
year changing more quickly in more recent years at Lake Sunapee?"

**Answer 11:**

Use a mix of code, figures, and text to present your analysis.

# Are lakes across the globe showing similar patterns to Lake Sunapee

To answer this question, you need to use a set of lakes in the doy-off
dataset.  We are going to focus on the following lakes:  `Baikal`, `Cazenovia`, `Mendota`,`Monona`, `Oneida`, `Sunapee` `Wingra`.  Use the `%in%` function to filter the dataset.  Learn more about `%in%` [here](https://www.datasciencemadesimple.com/in-operator-in-r/)

**Question 12:**

Present an analysis evaluates how the rate of change in doy-off day-of-year is
changing over time across the globe (i.e., the seven lakes listed above). Questions you should address in your
analysis

-   What is the distribution of the slopes for the different lake? Are they all
    the same sign?
-   What lake os showing the fastest rate of change in ice cover?
-   Do any of the following characteristics relate to differences in the slope?

```{r}
lake_characteristics <- tribble(~lake_name, ~latitude, ~lake_area, ~tropic_status,
             "Baikal", 53, 31722, "Oligotrophic",
             "Cazenovia",  42, 4.5, "Eutrophic",
             "Mendota",  43, 39, "Eutrophic",
             "Monona",  43, 13, "Eutrophic",
             "Oneida", 43, 207, "Mesotrophic",
             "Sunapee",  43, 17, "Oligotrophic",
             "Wingra",  43, 1, "Eutrophic")
```             
**Answer 12:**

# Citation

This module was initially developed as a Excel-based activity by Carey, C.C.,
J.L. Klug, and D.C. Richardson. 1 April 2015. Project EDDIE: Lake Ice Phenology.
Project EDDIE Module 1, Version 1:
<http://cemast.illinoisstate.edu/data-for-students/modules/ice-phenology.shtml>.
Module development was supported by NSF DEB 1245707.
